Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-09 manuscript/manuscript.Rmd
updates ms with new figure 1

*Andrew T. Tredennick ([atredenn@gmail.com](mailto:atredenn@gmail.com)), Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

@@

@@

Abstract
--------

*Keywords*: forecasting, climate change, grassland, integral projection model

@@

Here, we test the assumption that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data. We use a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments. We first fit population models with interannual variation in vital rates explained, in part, by climate covariates. We then perturb the climate covariates to test the sensitivities of species to climate change. By doing these analyses using both individual and aggregated forms of the same data, we provide a rigorous test of our hypothesis. We find that...

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therein for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

@@

In this paper, we model populations based on two levels of data: individual and quadrat. The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

Figures
-------

>>>>

*Andrew T. Tredennick[^corrauth], Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

@@

[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)

@@

Abstract
--------
The ability of population models to skillfully forecast future states under climate change is constrained by the limited spatial and temporal extent of demographic data. This data is often limited because it is difficult and costly to collect. An alternative is to rely on aggregate, population-level data that is easier and less costly to collect. Doing so requires assuming that population-level data accurately represents the aggregate response of the individuals that actually respond to weather. We tested this assumption using population models of four Montana grassland species fit using individual and aggregated forms of the same data. We fit population models with interannual variation in vital rates explained, in part, by climate covariates and then perturbed observed climate to compare model forecasts. Population models based on individual-level demographic data outperformed the models based on aggregate-level data in terms of accuracy and precision. The two model types produced inconsistent forecasts when we perturbed climate. Thus, it seems that, at least for the species in this location, demographic data is necessary to pick up key climate drivers of vital rates like survival that are not well resolved in aggregated data. On a pessimistic note, our work shows that even with detailed demographic data, model forecasts are extremely uncertain. It is becoming clear the long time series (over two decades) are necessary to produce meaningful forecasts.

*Keywords*: forecasting, climate change, grassland, integral projection model, population model

@@

Here, we test the assumption that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data. We use a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments. We first fit population models with interannual variation in vital rates explained, in part, by climate covariates. We then perturb the climate covariates to test the sensitivities of species to climate change. By doing these analyses using both individual and aggregated forms of the same data we can directly compare the two types of models.

In general, we find that population models based on detailed demographic data are more accurate and precise than models based on aggregated data. Both types of models are able to detect climate signals, as evidenced by the sensitivity of simulated equilbrium plant cover under a perturbed climate scenario. But the two types of models produce inconsistent forecasts, in some cases producing completely opposing predictions. This leads us to conclude that, at least for these species at this location, detailed demographic data is necessary to detect the "right" climate signal. A worrying caveat to our work is that forecasts from both models were very uncertain. It seems that even 14 years worth of demographic data is not enough to produce meaningful forecasts when model uncertainty is explicitly considered.

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

[^link]: http://esapubs.org/archive/ecol/E092/143/

@@

In this paper, we model populations based on two levels of data: individual and quadrat (Figure 1). The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

Figures
-------
```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.width=5, fig.height=8}
library(png)
library(grid)
img <- readPNG("components/figure/micromeso_flow.png")
grid.raster(img)
```

### 2015-06-09 manuscript/manuscript.tex
updates ms with new figure 1

\emph{Andrew T. Tredennick
(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT}

@@

\emph{Keywords}: forecasting, climate change, grassland, integral
projection model

@@

Here, we test the assumption that statistical and population models
based on aggregated data can detect climate signals as wells as models
based on individual-level data. We use a unique demographic dataset that
tracks the fates of individual plants from four species over 14 years to
build single-species population models, since those are often used tools
for ecological forecasts and climate vulnerability assessments. We first
fit population models with interannual variation in vital rates
explained, in part, by climate covariates. We then perturb the climate
covariates to test the sensitivities of species to climate change. By
doing these analyses using both individual and aggregated forms of the
same data, we provide a rigorous test of our hypothesis. We find
that\ldots{}

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives (Anderson et al.
2011), and interested readers should refer to the metadata therein for a
complete description. The site is about 800 m above sea level and mean
annual precipitation (1878-2009) is 334 mm, with most annual
precipitation falling from April through September. The site is grass
dominated and, for the purposes of our study, we focus on the four most
abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
\emph{Poa secunda} (POSE).

@@

In this paper, we model populations based on two levels of data:
individual and quadrat. The individual data is the ``raw'' data. For the
quadrat level we data we simply sum individual areal cover for each
quadrat by species. This is equivalent to a perfect census of quadrat
percent cover, so we do not need to consider measurement error. Based on
these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

@@

>>>>

\emph{Andrew T. Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}},
Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT}

@@

The ability of population models to skillfully forecast future states
under climate change is constrained by the limited spatial and temporal
extent of demographic data. This data is often limited because it is
difficult and costly to collect. An alternative is to rely on aggregate,
population-level data that is easier and less costly to collect. Doing
so requires assuming that population-level data accurately represents
the aggregate response of the individuals that actually respond to
weather. We tested this assumption using population models of four
Montana grassland species fit using individual and aggregated forms of
the same data. We fit population models with interannual variation in
vital rates explained, in part, by climate covariates and then perturbed
observed climate to compare model forecasts. Population models based on
individual-level demographic data outperformed the models based on
aggregate-level data in terms of accuracy and precision. The two model
types produced inconsistent forecasts when we perturbed climate. Thus,
it seems that, at least for the species in this location, demographic
data is necessary to pick up key climate drivers of vital rates like
survival that are not well resolved in aggregated data. On a pessimistic
note, our work shows that even with detailed demographic data, model
forecasts are extremely uncertain. It is becoming clear the long time
series (over two decades) are necessary to produce meaningful forecasts.

\emph{Keywords}: forecasting, climate change, grassland, integral
projection model, population model

@@

Here, we test the assumption that statistical and population models
based on aggregated data can detect climate signals as wells as models
based on individual-level data. We use a unique demographic dataset that
tracks the fates of individual plants from four species over 14 years to
build single-species population models, since those are often used tools
for ecological forecasts and climate vulnerability assessments. We first
fit population models with interannual variation in vital rates
explained, in part, by climate covariates. We then perturb the climate
covariates to test the sensitivities of species to climate change. By
doing these analyses using both individual and aggregated forms of the
same data we can directly compare the two types of models.

In general, we find that population models based on detailed demographic
data are more accurate and precise than models based on aggregated data.
Both types of models are able to detect climate signals, as evidenced by
the sensitivity of simulated equilbrium plant cover under a perturbed
climate scenario. But the two types of models produce inconsistent
forecasts, in some cases producing completely opposing predictions. This
leads us to conclude that, at least for these species at this location,
detailed demographic data is necessary to detect the ``right'' climate
signal. A worrying caveat to our work is that forecasts from both models
were very uncertain. It seems that even 14 years worth of demographic
data is not enough to produce meaningful forecasts when model
uncertainty is explicitly considered.

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
(Anderson et al. 2011) , and interested readers should refer to the
metadata therein for a complete description. The site is about 800 m
above sea level and mean annual precipitation (1878-2009) is 334 mm,
with most annual precipitation falling from April through September. The
site is grass dominated and, for the purposes of our study, we focus on
the four most abundant graminoid species: \emph{Bouteloua gracilis}
(BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
(PASM), and \emph{Poa secunda} (POSE).

@@

In this paper, we model populations based on two levels of data:
individual and quadrat (Figure 1). The individual data is the ``raw''
data. For the quadrat level we data we simply sum individual areal cover
for each quadrat by species. This is equivalent to a perfect census of
quadrat percent cover, so we do not need to consider measurement error.
Based on these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

\subsubsection{Vital rate sensitivities to climate
(IPM)}\label{vital-rate-sensitivities-to-climate-ipm}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-unnamed-chunk-1.pdf}
\caption{Work flow of the data aggregation, model fitting, and
population simulating.}
\end{figure}
